The role of standing variation in geographic convergent adaptation

### 2015-04-20 response-amnat-reviews.tex

After 90: ``somewhat larger probability of local establishment'' -- it is not at all clear to me what local establishment means in the case of a deleterious allele. Can you unpack this please?

\response{We meant establishment following the environmental shift, we have altered the text to reflect that.}

Equation 2: If the mutation selection balance is $\mu/s_d$, then the number of copies of the allele on average per unit area is $\rho mu/ s_d$. The probability that this leads to a successfully establishment is $2 s_b$ per copy approximately, which gives $2 s_b \mu \rho / s_d$ for $\lambda_0$. This seems like an easier derivation to me. (For large $s_d$, this diverges from your result -- which is more correct?)

\response{This difference arises from taking a continuous time approach. We have extended the description here to more clearly give the ``classic'' mutation-selection balance interpretation. We think it is important to work through this in the simple spatial case (via branching processes) rather than just move straight from the result people are familiar with deriving under the panmictic model. This allows folks to see that the result is valid in our spatial setting. }

%\gc{Peter is the difference coming about in part because the deterministic result is an approximation? Or because we are working in continuous time (or another reason?)?}
%\plr{Continuous time: in a discrete-time model, the mean density of individuals resulting from a rate-1 Poisson rain of branching processes would be $1/s_d$; in continuous time, it would be $1/\log(1/(1-s_d))$.  So, I wouldn't say that one is more valid that the other.  Hmph, confusing.}

@@

$$Ncrit == (\sigma * \log[1-s]) / (\sqrt{2}  \mu  \pi * (\sqrt{s} \log[1-s] - \sqrt{2}\sigma) )$$

(see attached mathematica file for a more clear version). If N is less than this value, then the characteristic length would be greater than 1, and convergent genetic evolution would not be expected. I'm sure there are some nicer ways that this could be expressed to show various features of importance when certain parameters are small (for example, when sigma is small, it drops out of the denominator, and then Ncrit is a simple function of sigma in the numerator). The scale of sigma is now expressed as a proportion of the total species range, so sigma can realistically be quite small for some species. Of course, this doesn't help us know how fast the characteristic length decreases once $N > Ncrit$, but Eq. 4 does that. I'm not sure if I've made an obvious blunder in here somewhere, but perhaps this is a useful approach? In any case, it seems worth acknowledging in the discussion that some species are quite unlikely to have convergent genetic evolution, due to their small range size, total population size, or high dispersal distance. The discussion currently says that convergent genetic evolution may be quite common (and line 466 even suggests it may be rare for gene flow to spread a selected allele across the range of a species). Delineating when this is important is especially interesting because the signal of convergent evolution can resemble local adaptation (e.g., line 400-408). If this approach makes sense, it could be discussed along with the section from lines 225-229, which presents a very important conclusion arising from this work.

\response{Thanks, this is a nice way of interpreting the result! We have added a paragraph giving this intepretation in roughly the locaiton in the text suggested.}

%\gc{

%Not sure this helps much. Thoughts Peter? By equating $\rho$ to N is he implicitly measuring $\sigma$ in units of range radius?}
%\plr{I like it.  (and, he mentions that $\sigma$ is as a fraction of the range, er, should be range diameter, I think.)
%What this comes down to is $N > \frac{ \sigma }{ \sqrt{2 s_b} \mu \pi ( 1 + \sigma \sqrt{s_b}/\log(1/(1-s_d)))}$,
%which since $\pi \sqrt{2s_b}$ will be something of order 1, is roughly just $N > \sigma/\mu$, or $N \mu > \sigma$.
%This is a bit surprising, but very concrete.  What do you think?
%}

@@

\response{We are interested in this area, and agree with the reviewer, but we don't want to expand on it given that the paper is already heavy on discussion.  In general we think that highly polygenic traits may often result in very different sets of alleles/loci being used in different populations. We are currently thinking on expanding on this point to form another paper. }
%\plr{er, should say a bit more here, but unsure how to say something that's not ``yes, we know.'' Isn't there more discussion of this in RC10?}

### 2015-04-20 standing-variation-paper.tex
tidied up .bib
Graham's final changes

\documentclass{article}
\usepackage{fullpage}
\usepackage{amsmath,amssymb}
\usepackage[authoryear]{natbib}
\usepackage{marginnote}
\usepackage{graphicx}
\usepackage{hyperref}
%\usepackage{latexml} % for \iflatexml
\usepackage{color}
\usepackage[dvipsnames]{xcolor}
\usepackage[utf8]{inputenc}  % for special characters in authors' names

\newcommand{\gc}[1]{{\it\color{blue}#1}}
\newcommand{\mfp}[1]{{\it\color{red}#1}}
\newcommand{\plr}[1]{{\it\color{Fuchsia}#1}}

\usepackage{lineno}  %%%% FOR AMNAT
\newcommand{\var}{\mathop{\mbox{Var}}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\R}{\mathbb{R}}

@@

%\iflatexml  % no margin notes in latexml
% \newcommand{\marginnote}[1]{}
%\newcommand{\marginnote}[1]{{\it\color{red}(#1)}}
%\fi

@@

@@

We can also find the mean proportion of space covered by standing
variants (we will restrict ourselves to two dimensions).
At time $t$ a geographic location has not yet been reached by the
mutation with probability given by equation \eqref{eqn-no-mut-t}.
\mfp{Given that it has not been reached by $t$, the probability} that it will be reached by time $t+dt$ 
by a standing variant is approximately $2 \lambda_0 \pi v^2 t dt$,
which is $\lambda_0$ multiplied by the thin slice of extra area in our
expanded circle at $t=0$, which has gone from a radius $vt$ to
$v(t+dt)$. The corresponding probability that the point is reached by a new
variant is 
$\lambda \pi v^2 t^2 dt$, which is $\lambda$ multiplied by the
sliver of  extra volume in our space-time cone at time $t+dt$ compared to that at time $t$.
Therefore, the mean proportion of space covered by standing variants,
is
\begin{equation} \label{prop_space_standing}
    z_0 = \int_0^\infty {2 \lambda_0 \pi v^2 t} \; \exp \left( -
      \lambda_0 \pi v^2 t^2 - \lambda \pi v^2 t^3 / 3 \right) dt,
\end{equation}
\mfp{this is the probability a given location is reached first by a
standing variant (which follows from competing our two exponential
waiting times).}
\gc{For applications we evaluate this integral numerically.}
%To evaluate this integral, again see Appendix \ref{apx:integrals}.
%\marginnote{put in this value in terms of Gaussian CDF?}

>>>>

\documentclass{article}
\usepackage{fullpage}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[authoryear]{natbib}
\usepackage{graphicx}
\usepackage{hyperref}
%\usepackage{latexml} % for \iflatexml
\usepackage{color}
\usepackage[dvipsnames]{xcolor}
\usepackage[utf8]{inputenc}  % for special characters in authors' names

\newcommand{\gc}[1]{{\it\color{blue}#1}}
\newcommand{\mfp}[1]{{\it\color{red}#1}}
\newcommand{\plr}[1]{{\it\color{blue}#1}}

\usepackage{lineno}  %%%% FOR AMNAT
\newcommand{\var}{\mathop{\mbox{Var}}}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\E}{\mathbb{E}}

@@

@@

\gc{Finally, while the form of eqn. \eqref{eqn:defines_chi}, specifying
the characteristic length, is not
particularly intuitive we can use it to ask when we should expect
multiple adaptative alleles within a large geographic region  where our selective pressure
is present (thanks to Sam Yeaman for proposing this
interpretation). Consider the case where this geographic area is a fairly regular
shape of area $G$ and diameter $\sqrt{G}$. Denote the total effective population size
over this area by $N = \rho G$, and the standard deviation of dispersal as a
fraction of the diameter of this area by $\sigma_G  = \sigma/
\sqrt{G}$. Measuring distance in units of the diameter of $\sqrt{G}$,
we expect multiple mutations when $\chi<1$ a
condition which will be met when 
\begin{equation}
N > \frac{\sigma_G}{ \mu \pi \sqrt{2s_b} \left(1 -  \sigma_G \sqrt{2s_b} /\log (1-s_d) \right)},
\end{equation}
this is found by rearranging eqn. \eqref{eqn:defines_chi} having set
$\chi=1$. This nicely shows that $\sigma_G/\mu$ is ae primary
determinant of the critical population density necessary for
convergent evolution.}

@@

We can also find the mean proportion of space covered by standing
variants (we will restrict ourselves to two dimensions).
At time $t$ a geographic location has not yet been reached by the
mutation with probability given by equation \eqref{eqn-no-mut-t}.
\mfp{Given that it has not been reached by $t$, the probability} that it will be reached by time $t+dt$ 
by a standing variant is approximately $2 \lambda_0 \pi v^2 t dt$,
which is $\lambda_0$ multiplied by the thin slice of extra area in our
expanded circle at $t=0$, which has gone from a radius $vt$ to
$v(t+dt)$. The corresponding probability that the point is reached by a new
variant is 
$\lambda \pi v^2 t^2 dt$, which is $\lambda$ multiplied by the
sliver of  extra volume in our space-time cone at time $t+dt$ compared to that at time $t$.
Therefore, the mean proportion of space covered by standing variants,
is
\begin{equation} \label{prop_space_standing}
    z_0 = \int_0^\infty {2 \lambda_0 \pi v^2 t} \; \exp \left( -
      \lambda_0 \pi v^2 t^2 - \lambda \pi v^2 t^3 / 3 \right) dt,
\end{equation}
\mfp{this is the probability a given location is reached first by a
standing variant (which follows from competing our two exponential
waiting times).}
\gc{For applications we evaluate this integral numerically.}
%To evaluate this integral, again see Appendix \ref{apx:integrals}.
%\plr{put in this value in terms of Gaussian CDF?}
